---
title: "Metagenomic ORF Detection and Annotation"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = FALSE)
options(scipen=999)
cpal=4
```

## ORF detection

```{r, echo=FALSE}
sfile=read.table(commandArgs(trailingOnly=T)[1],sep="\t",header=T)
norfs=nrow(sfile)
nconts=length(unique(sfile$Contig))
orflen = format(round(mean(sfile$Size)),big.mark=",")
orflensd = format(round(sd(sfile$Size)),big.mark=",")

orfspercont=data.frame(table(sfile$Contig))
meanopc=round(mean(orfspercont$Freq))
sdopc=round(sd(orfspercont$Freq))

```

A total of **`r norfs`** ORFs were detected by running Prodigal on **`r nconts`** input contigs.

There was a mean of **`r meanopc` $\pm$ `r sdopc`** ORFs per contig.

```{r, echo=FALSE}
library(ggplot2)
library(RColorBrewer)
orfspercontplot = ggplot(orfspercont,aes(Freq))+geom_histogram(bins = 50,fill=brewer.pal(3,"Accent")[1])+theme_classic()+ylab("Count")+xlab("No. Contigs per ORF")+guides(fill=FALSE)
plot(orfspercontplot)
```


ORFs had a mean length of **`r orflen` $\pm$ `r orflensd`** (Mean$\pm$SD).

```{r, echo=FALSE}
library(ggplot2)
library(RColorBrewer)
orflenhist = ggplot(sfile,aes(Size))+geom_histogram(bins = 50,fill=brewer.pal(3,"Accent")[2])+theme_classic()+ylab("Count")+xlab("Contig Size")+guides(fill=FALSE)
plot(orflenhist)
```
 
## Functional Annotation
 
```{r, echo=FALSE}
noannot=length(which(is.na((sfile$No_GO_Terms))))
perannot=round((1-(noannot/nrow(sfile)))*100)
meango = format(mean(sfile$No_GO_Terms,na.rm=T),digits = 3)
meankeg = format(mean(sfile$No_KEGG_KOs,na.rm=T),digits = 3)
meanogs = format(mean(sfile$No_OGs,na.rm=T),digits = 3)
meancogs = format(mean(sfile$No_COG_Cats,na.rm=T),digits=3)


sdgo = format(sd(sfile$No_GO_Terms,na.rm=T),digits = 3)
sdkeg = format(sd(sfile$No_KEGG_KOs,na.rm=T),digits = 3)
sdogs = format(sd(sfile$No_OGs,na.rm=T),digits = 3)
sdcogs = format(sd(sfile$No_COG_Cats,na.rm=T),digits=3)

pergo=round((length(which(sfile$No_GO_Terms>0))/nrow(sfile))*100)
perko=round((length(which(sfile$No_KEGG_KOs>0))/nrow(sfile))*100)
perogs=round((length(which(sfile$No_OGs>0))/nrow(sfile))*100)
percogs=round((length(which(sfile$No_COG_Cats>0))/nrow(sfile))*100)
```

**`r perannot`%** of ORFs were assigned at least one functional annotation using EggNog-Mapper.

```{r, echo=FALSE}
library(knitr)
formed=data.frame(cbind(c(meango,meankeg,meanogs,meancogs),c(sdgo,sdkeg,sdogs,sdcogs),c(pergo,perko,perogs,percogs)))
rownames(formed)=c("GO Terms","KEGG KOs","EggNOG OGs","COG Categories")
colnames(formed)=c("Mean Number per ORF","SD Number per ORF","Percent ORFs with at Least One")
kable(formed,caption="Functional annotation summaries per annotation type.")

```

## Taxonomic Annotation

```{r, echo=FALSE}
nospec=which(is.na(sfile$species))
taxper=round(100-((length(nospec)/nrow(sfile))*100))
```
 
 **`r taxper`%** of ORFs were assigned taxonomy at some level using DIAMOND against the NCBI nr database and the LCA algorithm in MEGAN.
 
 The following plots use a looping colour pallete. Cases where the same colour occurs multiple times within a bar represent distinct taxa.
 
 **Distribution of taxa across all ORFs**
 
```{r, echo=FALSE,message=FALSE,warning=FALSE}
justtax=sfile[-nospec,]
taxsplit=strsplit(as.character(justtax$species),"|",fixed=TRUE)
taxsplit=data.frame(matrix(unlist(taxsplit),nrow=nrow(justtax),byrow=T))
taxsplit=cbind(as.character(justtax$ORF),taxsplit)
colnames(taxsplit)=c("ORF","Domain","Phyla","Class","Order","Family","Genus","Species")

library(reshape2)
library(cowplot)
taxdat=melt(taxsplit, id="ORF")
cols=rep(brewer.pal(12,"Set3"),100)
plotlayout=list(geom_bar(),coord_flip(),theme_classic(),guides(fill=FALSE),scale_fill_manual(values=c(cols)),
                theme(axis.text.y =element_blank(),axis.ticks.y=element_blank(),axis.line.y =element_blank(),axis.title.y=element_text(size=8)))
dp = ggplot(taxdat[which(taxdat$variable=="Domain"),],aes(x=1,fill=value))+plotlayout+xlab("Domain")+ylab("")
pp = ggplot(taxdat[which(taxdat$variable=="Phyla"),],aes(x=1,fill=value))+plotlayout+xlab("Phyla")+ylab("")
cp = ggplot(taxdat[which(taxdat$variable=="Class"),],aes(x=1,fill=value))+plotlayout+xlab("Class")+ylab("")
op = ggplot(taxdat[which(taxdat$variable=="Order"),],aes(x=1,fill=value))+plotlayout+xlab("Order")+ylab("")
fp = ggplot(taxdat[which(taxdat$variable=="Family"),],aes(x=1,fill=value))+plotlayout+xlab("Family")+ylab("")
gp = ggplot(taxdat[which(taxdat$variable=="Genus"),],aes(x=1,fill=value))+plotlayout+xlab("Genus")+ylab("")
sp = ggplot(taxdat[which(taxdat$variable=="Species"),],aes(x=1,fill=value))+plotlayout+xlab("Species")+ylab("ORF Count")

plot_grid(sp,gp,fp,op,cp,pp,dp,nrow=7)
  
```

  **Distribution of taxa per contig**
  
  Only considering ORFs assigned taxonomy.

```{r echo=FALSE}
tb=table(sfile$Contig)
sord=sfile
sord$Contig=factor(sord$Contig, levels = names(tb[order(tb, decreasing = TRUE)]))
sord=sord[-which(is.na(sord$species)),]
plotlayoutcont=list(geom_bar(),coord_flip(),theme_classic(),guides(fill=FALSE),scale_fill_manual(values=c(cols)),
                theme(axis.text.y =element_blank(),axis.ticks.y=element_blank(),axis.line.y =element_blank(),axis.title.y=element_text(size=12)))
spec=ggplot(sord,aes(x=Contig,fill=species))+geom_bar()+guides(fill=FALSE)+plotlayoutcont+ylab("No ORFs")
genc=ggplot(sord,aes(x=Contig,fill=genus))+geom_bar()+guides(fill=FALSE)+plotlayoutcont+ylab("No ORFs")
famc=ggplot(sord,aes(x=Contig,fill=family))+geom_bar()+guides(fill=FALSE)+plotlayoutcont+ylab("No ORFs")
ordc=ggplot(sord,aes(x=Contig,fill=order))+geom_bar()+guides(fill=FALSE)+plotlayoutcont+ylab("No ORFs")
clasc=ggplot(sord,aes(x=Contig,fill=class))+geom_bar()+guides(fill=FALSE)+plotlayoutcont+ylab("No ORFs")
phyc=ggplot(sord,aes(x=Contig,fill=phylum))+geom_bar()+guides(fill=FALSE)+plotlayoutcont+ylab("No ORFs")
domc=ggplot(sord,aes(x=Contig,fill=domain))+geom_bar()+guides(fill=FALSE)+plotlayoutcont+ylab("No ORFs")

sptab=table(sord$Contig,sord$species)
gntab=table(sord$Contig,sord$genus)
famtab=table(sord$Contig,sord$family)
ordtab=table(sord$Contig,sord$family)
clastab=table(sord$Contig,sord$class)
phytab=table(sord$Contig,sord$phylum)
domtab=table(sord$Contig,sord$domain)

```

**Species**

Each contig contained a mean of `r format(mean(rowSums(sptab!=0)),digits=2)` different species annotations.

The most common species annotations accounted for a mean of `r format(mean(apply(sptab,1,function(x)(max(x)/sum(x)*100))),digits=2)`% of the ORFs within a contig.

```{r echo=FALSE}
plot(spec)
```

**Genus**

Each contig contained a mean of `r format(mean(rowSums(gntab!=0)),digits=2)` different genus annotations.

The most common genus annotations accounted for a mean of `r format(mean(apply(gntab,1,function(x)(max(x)/sum(x)*100))),digits=2)`% of the ORFs within a contig.

```{r echo=FALSE}
plot(genc)
```

**Family**

Each contig contained a mean of `r format(mean(rowSums(famtab!=0)),digits=2)` different family annotations.

The most common family annotations accounted for a mean of `r format(mean(apply(famtab,1,function(x)(max(x)/sum(x)*100))),digits=2)`% of the ORFs within a contig.


```{r echo=FALSE}
plot(famc)
```

**Order**

Each contig contained a mean of `r format(mean(rowSums(ordtab!=0)),digits=2)` different order annotations.

The most common order annotations accounted for a mean of `r format(mean(apply(ordtab,1,function(x)(max(x)/sum(x)*100))),digits=2)`% of the ORFs within a contig.

```{r echo=FALSE}
plot(ordc)
```

**Class**

Each contig contained a mean of `r format(mean(rowSums(clastab!=0)),digits=2)` different class annotations.

The most common class annotations accounted for a mean of `r format(mean(apply(clastab,1,function(x)(max(x)/sum(x)*100))),digits=2)`% of the ORFs within a contig.

```{r echo=FALSE}
plot(clasc)
```

**Phylum**

Each contig contained a mean of `r format(mean(rowSums(phytab!=0)),digits=2)` different phylum annotations.

The most common phylum annotations accounted for a mean of `r format(mean(apply(phytab,1,function(x)(max(x)/sum(x)*100))),digits=2)`% of the ORFs within a contig.

```{r echo=FALSE}
plot(phyc)
```

**Domain**

Each contig contained a mean of `r format(mean(rowSums(domtab!=0)),digits=2)` different domain annotations.

The most common domain annotations accounted for a mean of `r format(mean(apply(domtab,1,function(x)(max(x)/sum(x)*100))),digits=2)`% of the ORFs within a contig.

```{r echo=FALSE}
plot(domc)
```


